100 years of Pies vs. Bars

### 2025-04-16 Draft.qmd
Added conclusion and correction

```{=tex}
\mt{They conducted a foundational study to assess how accurately people interpret different graphical encodings.
They focused on six basic graphical perception tasks: position along a common scale, position along non-aligned scales, length, angle, slope, and area (represented by circles and irregular "blobs").
Participants were shown 70 visual displays, each with one standard object and three comparisons, and asked to estimate each comparison as a percentage of the standard.
The study involved 127 participants---high school students, college students, and professionals---who each made 210 judgments across all task types and standard sizes.
Stimuli were randomized to avoid learning effects}
```
The main outcome was estimation accuracy---how close each response was to the true percentage.
No scales or measurement tools were provided, so judgments relied purely on visual perception.
Each of the six tasks was tested with both small and large reference standards, and values ranged from 17.5% to 87.5%.
Results revealed a clear perceptual hierarchy: position along a common scale led to the most accurate estimates, followed by length.
In contrast, angle and area judgments were significantly less accurate.
These findings have had a lasting impact on visualization design, encouraging the use of perceptually stronger encodings for quantitative data.}

@@

```{=tex}
\mt{Skau and Kosara (2016) conducted a focused experiment to isolate how individual visual encodings---arc length, angle, and area---contribute to interpretation accuracy in pie and donut charts.
They created six chart types: baseline pie, baseline donut, arc-only, angle (pie-style), angle (donut-style), and area-only.
Each chart showed just two segments---a blue "target" slice and a gray remainder---allowing participants to focus on estimating a single proportion without distractions.
Across 48 randomized trials (eight per chart type), 100 Mechanical Turk participants were asked the same question for every chart: "What percentage of the whole is indicated below?" The charts were randomly rotated to avoid alignment biases, and a brief tutorial was provided to ensure participants understood each chart forma}
```
Measurement & Accuracy The study focused exclusively on perceptual accuracy.
For each chart, researchers measured the absolute difference between the participant's estimate and the true value, as well as error patterns such as reversed judgments (e.g., answering 75% instead of 25% on angle charts).
No time constraints were enforced, and response time or subjective preferences were not formally recorded.
However, participants were asked before and after the main task which visual cue they believed they relied on---arc, angle, or area.
Results revealed that baseline pie charts led to the most accurate judgments overall, followed by arc-only encodings.
Surprisingly, many participants reported using angle judgments, even in chart types where angle cues were removed or degraded.}

@@

```{r}
#| echo: false

library(ggplot2)

# Data
chart_data <- data.frame(
  Chart_Type = rep(c("Bar", "Pie", "Donut"), each = 6),
  Task_Type = rep(c("A vs B", "A / (A + B)", "A vs (B + C)", 
                    "(A + B) vs (C + D)", "Ranking", "Magnitude Est."), 3),
  Rank = c(
    1, 3, 1, 2, 1, 3,  # Bar
    2, 1, 2, 1, 2, 1,  # Pie
    3, 2, 3, 3, 3, 1   # Donut
  )
)

# Reorder x-axis: Pie, Bar, Donut
chart_data$Chart_Type <- factor(chart_data$Chart_Type, levels = c("Pie", "Bar", "Donut"))

# Set Rank so that 1 shows on top
chart_data$Rank <- factor(chart_data$Rank, levels = c(3, 2, 1))

# Plot
ggplot(chart_data, aes(x = Chart_Type, y = Rank, fill = Task_Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(
    title = "Chart Type Performance Across Tasks",
    y = "Rank (1 = Best)",
    x = NULL,
    fill = "Task Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

@@

```{r}
#| echo: false

library(ggplot2)
library(dplyr)

# Data for 4 segments
df1 <- data.frame(
  group = c("A", "B", "C", "D"),
  value = c(25, 25, 25, 25)
)

ggplot(df1, aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity", color = "white") +
  coord_polar("y") +
  theme_void() +
  ggtitle("Comparison: A, B vs C, D") +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14)
  )

# Data for two pie charts
df2 <- data.frame(
  group = rep(c("A+B", "C+D"), each = 2),
  sub_group = rep(c("A", "B", "C", "D"), 1),
  value = c(20, 30, 25, 25)
)

df2$facet <- rep(c("Group AB", "Group CD"), each = 2)

ggplot(df2, aes(x = "", y = value, fill = sub_group)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y") +
  facet_wrap(~ facet) +
  theme_void() +
  ggtitle("Comparison: AB vs CD") +
  theme(
    strip.text = element_text(size = 12),
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14)
  )

```{r}
#| echo: False
library(ggplot2)
library(dplyr)

# Helper function to create pie chart data
make_pie_data <- function(labels, values, group) {
  df <- data.frame(
    label = labels,
    value = values
  )
  df$group <- group
  df
}

# Create all pie segments
pie_data <- bind_rows(
  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
)

# Prepare for pie chart layout
pie_data <- pie_data %>%
  group_by(group) %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = lag(ymax, default = 0),
    label_pos = (ymin + ymax) / 2
  )

# Plot
ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
  geom_rect(color = "white") +
  coord_polar(theta = "y") +
  facet_wrap(~group, ncol = 1, strip.position = "left") +
  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +
  scale_fill_manual(values = c("A" = "#e0f3db", "B" = "#ccebc5", "C" = "#b2df8a", "A+B" = "#a6cee3")) +
  theme_void() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  ggtitle("Comparison Tasks")

# ==== 1. Full Pie Chart: A, B, C, D ====
df_full <- data.frame(
  group = c("A", "B", "C", "D"),
  value = c(25, 25, 25, 25)
)

df_full <- df_full %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label_pos = (ymin + ymax) / 2
  )

ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
  geom_rect(color = "white") +
  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
  theme_void() +
  ggtitle("Full Pie Chart: A, B, C, D")

```{r}
#| echo: false
library(ggplot2)
library(dplyr)

# Helper function to create pie chart data
make_pie_data <- function(labels, values, group) {
  df <- data.frame(
    label = labels,
    value = values
  )
  df$group <- group
  df
}

# Combine all pie data
pie_data <- bind_rows(
  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
)

# Compute positions
pie_data <- pie_data %>%
  group_by(group) %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = lag(ymax, default = 0),
    label_pos = (ymin + ymax) / 2
  )

# Color palette (match original soft hues)
color_palette <- c("A" = "#FBB4AE", "B" = "#B3CDE3", "C" = "#DECBE4", "A+B" = "#CCEBC5")

# Plot
ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
  geom_rect(color = "white") +
  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +  # Labels inside
  coord_polar(theta = "y") +
  facet_wrap(~group, nrow = 1) +  # Horizontal layout
  scale_fill_manual(values = color_palette) +
  theme_void() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

```{r}
#| echo: false
# Load required libraries
library(ggplot2)
library(dplyr)

# ==== 1. Full Pie Chart: A, B, C, D ====
df_full <- data.frame(
  group = c("A", "B", "C", "D"),
  value = c(25, 25, 25, 25)
)

df_full <- df_full %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label_pos = (ymin + ymax) / 2
  )

ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
  geom_rect(color = "white") +
  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
  theme_void() +
  ggtitle("Full Pie Chart: A, B, C, D")

# ==== 2. Half Pie Chart: Group AB ====
df_ab <- data.frame(
  group = c("A", "B"),
  value = c(50, 50)
)

df_ab <- df_ab %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label_pos = (ymin + ymax) / 2
  )

p_ab <- ggplot(df_ab, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
  geom_rect(color = "white") +
  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
  coord_polar(theta = "y", start = pi) +  # start at top to show half-pie
  scale_fill_manual(values = gray.colors(2, start = 0.9, end = 0.7)) +
  theme_void() +
  ggtitle("Group AB")

# ==== 3. Half Pie Chart: Group CD ====
df_cd <- data.frame(
  group = c("C", "D"),
  value = c(50, 50)
)

df_cd <- df_cd %>%
  mutate(
    fraction = value / sum(value),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label_pos = (ymin + ymax) / 2
  )

p_cd <- ggplot(df_cd, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
  geom_rect(color = "white") +
  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
  coord_polar(theta = "y", start = pi) +
  scale_fill_manual(values = gray.colors(2, start = 0.6, end = 0.4)) +
  theme_void() +
  ggtitle("Group CD")

# ==== Combine AB and CD side by side ====
library(patchwork)
p_ab + p_cd
```

```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(patchwork)

# Function to make pie chart with highlighted segments
make_pie_plot <- function(data, title, highlight_group = NULL, highlight_color = "black") {
  data <- data %>%
    mutate(
      fraction = value / sum(value),
      ymax = cumsum(fraction),
      ymin = lag(ymax, default = 0),
      label_pos = (ymin + ymax) / 2,
      label = label
    )

  # Add border flag
  data$highlight <- ifelse(data$label %in% highlight_group, TRUE, FALSE)

  ggplot(data) +
    geom_rect(aes(
      ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
      fill = label, color = highlight
    ), size = 1) +
    coord_polar(theta = "y") +
    geom_text(aes(
      x = 1, y = label_pos, label = label
    ), size = 5) +
    scale_fill_manual(values = c(
      "A" = "#f4a582", "B" = "#92c5de", "C" = "#d1bce3", "A+B" = "#b2df8a", "B+C" = "#b2abd2"
    )) +
    scale_color_manual(values = c("TRUE" = highlight_color, "FALSE" = NA), guide = "none") +
    theme_void() +
    ggtitle(title) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Define the three datasets
data1 <- tibble(label = c("A+B", "C"), value = c(0.5, 0.5))  # (A+B) vs C
data2 <- tibble(label = c("A", "B", "C"), value = c(0.25, 0.25, 0.5))  # A vs (B+C)
data3 <- tibble(label = c("A", "B"), value = c(0.4, 0.6))  # A vs B

# Make each plot with outlines
p1 <- make_pie_plot(data1, "(A+B) vs C", highlight_group = "A+B")
p2 <- make_pie_plot(data2, "A vs (B+C)", highlight_group = c("B", "C"))
p3 <- make_pie_plot(data3, "A vs B", highlight_group = c("A", "B"))

# Combine the three plots horizontally
(p1 | p2 | p3)

@@

# Conclusion

@@

>>>>

\mt{They conducted a foundational study to assess how accurately people interpret different graphical encodings.
They focused on six basic graphical perception tasks: position along a common scale, position along non-aligned scales, length, angle, slope, and area (represented by circles and irregular "blobs").
Participants were shown 70 visual displays, each with one standard object and three comparisons, and asked to estimate each comparison as a percentage of the standard.
The study involved 127 participants---high school students, college students, and professionals---who each made 210 judgments across all task types and standard sizes.
Stimuli were randomized to avoid learning effects}

The main outcome was estimation accuracy---how close each response was to the true percentage.
No scales or measurement tools were provided, so judgments relied purely on visual perception.
Each of the six tasks was tested with both small and large reference standards, and values ranged from 17.5% to 87.5%.
Results revealed a clear perceptual hierarchy: position along a common scale led to the most accurate estimates, followed by length.
In contrast, angle and area judgments were significantly less accurate.
These findings have had a lasting impact on visualization design, encouraging the use of perceptually stronger encodings for quantitative data.}

@@

\mt{Skau and Kosara (2016) conducted a focused experiment to isolate how individual visual encodings---arc length, angle, and area---contribute to interpretation accuracy in pie and donut charts.
They created six chart types: baseline pie, baseline donut, arc-only, angle (pie-style), angle (donut-style), and area-only.
Each chart showed just two segments---a blue "target" slice and a gray remainder---allowing participants to focus on estimating a single proportion without distractions.
Across 48 randomized trials (eight per chart type), 100 Mechanical Turk participants were asked the same question for every chart: "What percentage of the whole is indicated below?" The charts were randomly rotated to avoid alignment biases, and a brief tutorial was provided to ensure participants understood each chart forma}

Measurement & Accuracy The study focused exclusively on perceptual accuracy.
For each chart, researchers measured the absolute difference between the participant's estimate and the true value, as well as error patterns such as reversed judgments (e.g., answering 75% instead of 25% on angle charts).
No time constraints were enforced, and response time or subjective preferences were not formally recorded.
However, participants were asked before and after the main task which visual cue they believed they relied on---arc, angle, or area.
Results revealed that baseline pie charts led to the most accurate judgments overall, followed by arc-only encodings.
Surprisingly, many participants reported using angle judgments, even in chart types where angle cues were removed or degraded.}

@@

@@

```{r}
# | echo: false
library(ggplot2)
library(dplyr)
library(patchwork)

# Helper function for pie chart
create_pie <- function(data, fill_colors, title, bold_groups = NULL) {
  data <- data %>%
    mutate(
      fraction = value / sum(value),
      ymax = cumsum(fraction),
      ymin = lag(ymax, default = 0),
      label_pos = (ymin + ymax) / 2,
      border_size = ifelse(label %in% bold_groups, 3, 0.8)
    )

  ggplot(data) +
    geom_rect(aes(ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
                  fill = label, size = border_size),
              color = "black", show.legend = FALSE) +
    geom_text(aes(x = 1, y = label_pos, label = label), size = 6) +
    scale_fill_manual(values = fill_colors) +
    scale_size_identity() +
    coord_polar(theta = "y") +
    labs(title = title) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
}

# Pie 1: A vs B
data1 <- tibble(label = c("B", "A"), value = c(0.35, 0.65))
fill1 <- c("A" = "#f4a582", "B" = "#92c5de")
p1 <- create_pie(data1, fill1, "A vs B")

# Pie 2: (A+B) vs C
data2 <- tibble(label = c("C", "B", "A"), value = c(0.25, 0.25, 0.5))
fill2 <- c("A" = "white", "B" = "#ff00ff", "C" = "#00ffff")
p2 <- create_pie(data2, fill2, "(A+B) vs C", bold_groups = c("A", "B"))

# Pie 3: (A+B) vs (C+D)
data3 <- tibble(label = c("C", "B", "A", "D"), value = c(0.25, 0.20, 0.35, 0.20))
fill3 <- c("A" = "white", "B" = "white", "C" = "#00ffff", "D" = "pink")
p3 <- create_pie(data3, fill3, "(A+B) vs (C+D)", bold_groups = c("A", "B"))

# Combine with patchwork
(p1 | p2 | p3)

@@

# Conclusion
As of now, no one has determined a definitive winner between pie charts and bar charts. These two charts remain essential in every part of science, business, sociology, and related fields. The efficacy of various charts is contingent upon ratio, measurement, and perception aspects of the representation. In research, a chart should not be dismissed solely because it is deemed inadequate.
Pie charts make it easier to assess proportions relative to the whole, but bar charts excel in comparing magnitudes and ranking activities. Performance fluctuates depending on whether the task entails value estimation, segment comparison, or identification of the major component. Generally, charts that clearly delineate comparative parts (e.g., adjacent slices or bold outlines) enhance interpretability.

We have also discovered an intriguing fact that, despite the evolution of formats and digital interfaces, fundamental perceptual concepts (@eellsRelativeMeritsCircles1926, @clevelandGraphicalPerceptionTheory1984 and so on) continue to have validity. Nonetheless, this finding does not signify the conclusion of the pursuit for the optimal chart. Future research can clarify the dispute through improved experimental design and more compelling methodologies.

@@

Numerous constraints are present in both early and contemporary investigations.  Initial trials (from the 1920s to 1980s) frequently exhibited insufficient statistical rigor, characterized by delicate measurement frameworks and rudimentary designs.  Contemporary research (@spenceDisplayingProportionsPercentages1991; @hill2025bar) has implemented enhanced experimental designs, however it continues to encounter difficulties, especially with participant diversity.

Contemporary research predominantly utilizes student samples, so constraining generalizability and potentially failing to represent wider perceptual disparities across age, education, or cultural contexts.  Subsequent investigations ought to tackle this issue by enhancing sample variety and refining experimental methodologies.  This may reveal fresh insights and promote more inclusive, contextually suitable chart type.

 Individuals' perceptions of charts may alter between devices—mobile phones, tablets, or desktop monitors—due to variations in screen size, resolution, and interaction methods.  Furthermore, enhancing the reliability of perceptual measurements is crucial.  Future research may include technologies such as eye-tracking, response time analysis, or cognitive load evaluations to provide more nuanced findings.  The objective is to convert experimental results into practical design advice that assist practitioners in choosing the most suitable chart kinds for certain audiences and jobs.
